### 2021-03-19 assertions/doc/thesis-bernhardt/chapters/1-introduction.tex
Offline Chapter beginnign
Started rework of Offline design

\chapter{Introduction}\label{chap:introduction}
Each execution of a blockchain-based Smart Contract in a gas-instrumented virtual machine generates costs proportional to the amount and complexity of computations and the  memory consumption \cite{goodman_tezos_2014}\cite{wood_ethereum_2021}. These fees, often calculated in gas, have to be paid for by the sender of the transaction and thus provides a powerful incentive to keep  smart contracts simple. However, in reality meaningful applications might need to perform some resource intensive computations at some point, which would be too expensive to execute inside the blockchain's virtual machine. Moreover, some blockchains impose a hard cap of how much gas can be consumed in one transaction, limiting the amount of operations that can be executed \cite{goodman_tezos_2014}.\\
In order to avoid these costs and restrictions, an application could perform some computations outside of the blockchain context (i.e., off-chain) and only submit the results to a smart contract as input parameter for further processing or to store it on the chain. This approach raises another problem though - trusting that the submitted results are correct poses a safety as well as a security risk. Invalid results can be submitted unintentionally, if the computation is faulty, or a third party may pass them intentionally to cause harm or take advantage. Consequently, submitted parameters need to be checked for specified properties before the smart contract is executed. Although in many cases, verification is less expensive than computing the result itself, the generated costs for checking the validity increase with the size of the problem or input, ultimately becoming too expensive. \\
Consider, for instance, the following contract from \cite{thiemann_2020} that expects a prime number :\todo{State programming language?!}
\begin{lstlisting}[caption=Smart contract expecting a prime number, numbers=none, language=Solidity, label=lst:prime]
contract Example {
	function ( int p ) public {
	// assume p is prime
	...
	}
}
\end{lstlisting}

@@

\begin{lstlisting}[caption=Smart Contract expecting a sorted array, numbers=none, language=Solidity]
contract Sorted {
  function find ( int [50] a , int v ) public {
    // assume a is sorted
  }
}
\end{lstlisting}

The assumption that the list is sorted, expressed in predicate logic, is
\begin{equation}\label{eq:sorted}
	(\forall n) (0 \leq n \le |a| - 1) \Rightarrow a[n] \leq a[n+1]
\end{equation}

@@

\begin{equation}\label{eq:sorted_neg}
	(\exists n) (0 \leq n \le |a| - 1) \wedge a[n] > a[n+1]
\end{equation}

@@

\move{maybe to detailed? move to efficiency section}

>>>>

\chapter{Introduction}\label{chap:introduction}
Each execution of a blockchain-based Smart Contract in a gas-instrumented virtual machine generates costs proportional to the amount and complexity of computations and the  memory consumption \cite{goodman_tezos_2014}\cite{wood_ethereum_2021}. These fees, often calculated in gas, have to be paid for by the sender of the transaction and thus provides a powerful incentive to keep  smart contracts simple. However, in reality meaningful applications might need to perform some resource intensive computations at some point, which would be too expensive to execute inside the blockchain's virtual machine. Moreover, some blockchains impose a hard cap of how much gas can be consumed in one transaction, limiting the amount of operations that can be executed \cite{goodman_tezos_2014}.\\
In order to avoid these costs and restrictions, an application could perform some computations outside of the blockchain context (i.e., off-chain) and only submit the results to a smart contract as input parameter for further processing or to store it on the chain. This approach raises another problem though - trusting that the submitted results are correct poses a safety as well as a security risk. Invalid results can be submitted unintentionally, if the computation is faulty, or a third party may pass them intentionally to cause harm or take advantage. Consequently, submitted parameters need to be checked for specified properties before the smart contract is executed. Although in many cases, verification is less expensive than computing the result itself, the generated costs for checking the validity increase with the size of the problem or input, ultimately becoming too expensive. \\
Consider, for instance, the following contract from \cite{thiemann_2020} that expects a prime number: \todo{State programming language?!}
\begin{lstlisting}[caption=Smart contract expecting a prime number, numbers=none, language=Solidity, label=lst:prime]
contract Example {
	function ( int p ) public {
	// assume p is prime
	...
	}
}
\end{lstlisting}

@@

\begin{lstlisting}[caption=Smart Contract expecting a sorted array, numbers=none, language=Solidity, label=lst:sorted]
contract Sorted {
  function find ( int [50] a , int v ) public {
    // assume a is sorted
  }
}
\end{lstlisting}

The assumption that the list is sorted, expressed in predicate logic, is
\begin{equation}\label{eq:sorted}
	(\forall n) (0 \leq n < |a| - 1) \Rightarrow a[n] \leq a[n+1]
\end{equation}

@@

\begin{equation}\label{eq:sorted_neg}
	(\exists n) (0 \leq n < |a| - 1) \wedge a[n] > a[n+1]
\end{equation}

@@

\move{maybe too detailed? move to efficiency section}

### 2021-03-19 assertions/doc/thesis-bernhardt/chapters/2-offline_design.tex
Started rework of Offline design

\chapter{Offline Design}\label{chap:offline}
Implementing distributed assertion checking on Tezos requires some off-chain considerations and infrastructure. The previous chapter defined a formal set of logical formulae amenable for this approach and gave some in-depth examples. \todo{Allgemeine Beschreibung was das Ziel des Chapters ist -> wir müssen aus einer Formula nun einen Contract erstellen, der einen assertion check ausführen kann. -> Define a syntax and a compilation pipeline that outputs a Michelson sc that can be originated.}

A concrete syntax for writing assertions on these formulae is described in the subsequent section. It will also show how some of the previously given examples are expressed in that syntax. The transformation of the original assertions to assertions that check for counterexamples or proofs is formalized in \secref{sec:transformation}. \todo{rest}

\section{Assertion syntax}

Argument against if/else conditions outside of assert():
In order to take advantage of this, the assertion syntax would need to be able to create different generators for different conditions. Currently, we only allow \texttt{if-else} instructions inside an \texttt{assertion} expression; this use-case could be an argument to lift that restriction and allow contracts like this:
%\lstinputlisting[caption=Assertion contract with conditional generators, label={lst:coprime}]{listings/coprime.tza}

However, this will make the transformation more involved, as the bounds cannot simply be merged with the generator anymore.

@@

\subsection{Defining the domain of random generators}
In order to assign each explicit bound to the respective quantifier, the formula has be skimmed for atomic constraints that contain bound variables. They're then moved and assigned to the quantifier that bounds the variable. If the constraint contains more than one, it is assigned to the quantifier with the highest depth in the order of quantifiers. \\
Consider a variation of the example given in \eqref{eq:sorted}, a formula checking if a given list is sorted:
\begin{equation}\label{eq:sorted_v2}
	(\forall n : int)(\forall m : int) (0 \leq n \le m < |a|) \Rightarrow a[n] \leq a[m]
\end{equation}
The premise can be considered as a conjunction of the four constraints $0 \leq n$, $n < |a| $, $n \le m$ and $m < |a|$. Applying the rules given above after negating the formula, the constraints are assigned to the quantifiers as follows:
\begin{equation}\label{eq:sorted_v2_bounds}
	(\exists n : int, 0 \leq n \wedge n < |a|)(\exists m : int, m > n \wedge m < |a|) \text{ } a[n] > a[m]
\end{equation}
Ultimately, the quantifiers are then translated to the corresponding random generators shown in \lstref{lst:rand}.
\begin{lstlisting}[label=lst:rand]
n = random(0, size(list))
m = random(n, size(list))
\end{lstlisting}

@@

\subsection{Interesting formulae}\label{sec:example_formulae}
- paper

- First of all some offline stuff to think about - formalize which formulas are amenable for this checking. Formalize what to do for which formulas (exists/forall/mixed) etc.
- Grammar/language to express these formulas (still general, not protocol specific)
	-> which operations do we need?
	-> use the form of expose?
	-> uses ocaml/michelson like syntax; used by other BC as well though
- effectiveness
- formalize the costs for checking these formulas

>>>>

\externaldocument{1-introduction}

\chapter{Generic Offline Design}\label{chap:offline}
The implementation of distributed assertion checking on any blockchain requires some off-chain considerations and infrastructure. The previous chapter defined a formal set of logical formulae amenable for this approach and gave some in-depth examples. As was stated before, the proposed implementation in the following chapters is, for the most part, restricted to formulas of predicate logic using universal quantification.\\
The purpose of the offline infrastructure is to provide a syntax for stating assertions that check logical formulas, as well as a toolchain to compile these assertions into code that can be originated on the target blockchain. Depending on the individual solution for the respective blockchain, this can either be the original contract extended with the assertion code, or separate code that extends the contract in some other way. This chapter describes the generic part of the offline design, which corresponds to the front-end of the compilation pipeline shown in \figref{fig:pipeline_generic}. As a start, \secref{sec:syntax} describes the concrete syntax for writing assertions and shows how it expresses some of the previously given examples. The transformation of the original assertions to assertions that check for counterexamples is formalized in \secref{sec:transformation} \todo{how to handle proofs}. Lastly, \secref{sec:effectiveness} analyses the reliability of distributed assertion checking.

\section{Assertion syntax}
The assertion syntax is described in two versions by the EBNF grammars shown in appendix \ref{apx:grammar}. One version implements a prefix and the other one an infix notation. The pipeline currently supports the prefix notation with mandatory parentheses to avoid the need of handling operator precedences. To avoid context switches (at least for developers on Tezos), the assertion syntax references those of OCaml and Michelson.\\
A file containing assertions (henceforth called `assertion contract') contains at least one assertion for some function or entrypoint of the original contract (henceforth called `parent contract'). An assertion begins with a signature consisting of an optional tag followed by a parameter type declaration. The tag can be used for documentation, but may sometimes be needed to indicate to which function of the parent contract the assertion is assigned. The body is an (optional) nesting of quantifiers and conditions around exactly one assertion. Conditionals can be used to restrict the quantification domains or for some other constraints. For completeness, the existential quantifier is already included in the grammar, however the current version of the pipeline will reject any assertions containing it.\\
Consider a variation of the formula in predicate logic that checks whether a given list is sorted in ascending order:
\begin{equation}\label{eq:sorted_v2}
	(\forall n : int)(\forall m : int) (0 \leq n < m < |a|) \Rightarrow a[n] \leq a[m]
\end{equation}
A respective assertion that checks this property for the parent contract from \lstref{lst:sorted} could be expressed as follows:
\begin{lstlisting}[language=Assertion]
(entrypoint %find (a : (list int))
  (forall (n : int)
    (forall (m : int)
      (if (le 0 n)
        (if (lt n m)
          (if (lt n (size a))
            (assert (le (nth a n) (nth a m)))))))))
\end{lstlisting}

\subsection{Extensions}
% possible extensions: variables, user defined functions

%Argument against if/else conditions outside of assert():
%In order to take advantage of this, the assertion syntax would need to be able to create different generators for different conditions. Currently, we only allow \texttt{if-else} instructions inside an \texttt{assertion} expression; this use-case could be an argument to lift that restriction and allow contracts like this:
%\lstinputlisting[caption=Assertion contract with conditional generators, label={lst:coprime}]{listings/coprime.tza}
%However, this will make the transformation more involved, as the bounds cannot simply be merged with the generator anymore.

@@

\subsection{Defining the domain of random generators}
In order to assign each explicit bound to the respective quantifier, the formula has be skimmed for atomic constraints that contain bound variables. They're then moved and assigned to the quantifier that bounds the variable. If the constraint contains more than one, it is assigned to the quantifier with the highest depth in the order of quantifiers. \\
Consider a variation of the example given in \eqref{eq:sorted}, a formula checking if a given list is sorted:
\begin{equation}\label{eq:sorted_v2}
	(\forall n : int)(\forall m : int) (0 \leq n \le m < |a|) \Rightarrow a[n] \leq a[m]
\end{equation}
\todo{variation given above. refer to that}

The premise can be considered as a conjunction of the four constraints $0 \leq n$, $n < |a| $, $n \le m$ and $m < |a|$. Applying the rules given above after negating the formula, the constraints are assigned to the quantifiers as follows:
\begin{equation}\label{eq:sorted_v2_bounds}
	(\exists n : int, 0 \leq n \wedge n < |a|)(\exists m : int, m > n \wedge m < |a|) \text{ } a[n] > a[m]
\end{equation}
Ultimately, the quantifiers are then translated to the corresponding random generators shown in \lstref{lst:rand}.
\begin{lstlisting}[label=lst:rand]
n = random(0, size(list))
m = random(n, size(list))
\end{lstlisting}

@@
